---
title: "Create ADSL"
author: "Christina Fillmore"
date: "03/02/2022"
output: html_document
description: "An example of how to generate an ADSL dataset from STDM"
---

## Introduction

This guide will show you how four pharmaverse packages, along with some from tidyverse, can be used to create an ADaM such as `ADSL` end-to-end, using the pilot CDISC SDTM data as input.

The four packages used with a brief description of their purpose are as follows:

* `{metacore}`: provides harmonized metadata/specifications object.
* `{metatools}`: uses the provided metadata to build/enhance and check the dataset.
* `{admiral}`: provides the ADaM derivations.
* `{xportr}`: delivers the SAS transport file (XPT) and eSub checks.

## Load Data and Required pharmaverse Packages

The first step is to load our pharmaverse packages and input data.

```{r setup, message=FALSE, warning=FALSE, results='hold'}
options(repos = c(
  pharmaverse = 'https://pharmaverse.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))

library(metacore)
library(metatools)
library(admiral.test)
library(admiral)
library(xportr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)

# Read in input SDTM data 
data("admiral_dm")
data("admiral_ex")
```

Next we need to load the specification file in the form of a `{metacore}` object.

```{r metacore, warning=FALSE, results='hold'}
# Read in metacore object 
load(metacore_example("pilot_ADaM.rda"))
metacore <- metacore %>% 
   select_dataset("ADSL")
```
  
## Start Building Derivations

The first derivation step we are going to do is to pull through all the columns that come directly from the SDTM datasets. You might know which datasets you are going to pull from directly already, but if you don't you can call `build_from_derived` with just an empty list and the error will tell you which datasets you need to supply. 

```{r, error=TRUE}
build_from_derived(metacore, list(), predecessor_only = FALSE)
```

In this case all the columns come from `DM` so that is the only dataset we will pass into `build_from_derived`. The resulting dataset has all the columns combined and any columns that needed renaming between SDTM and ADaM are renamed.

```{r demographcis}
adsl_preds <- build_from_derived(metacore, 
                                 ds_list = list("dm" = admiral_dm), 
                                 predecessor_only = FALSE, keep = TRUE)
head(adsl_preds, n=10)
```

Now we have the base dataset, we can start to create some variables. We can start with creating the subgroups using the controlled terminology, in this case `AGEGR1`. The metacore object holds all the metadata needed to make `ADSL`. Part of that metadata is the controlled terminology, which can help automate the creation of subgroups. We can look into the `{metacore}` object and see the controlled terminology for `AGEGR1`.

```{r}
get_control_term(metacore, variable = AGEGR1)
```

Because this controlled terminology is written in a fairly standard format we can automate the creation of `AGEGR1`. The `{metatools}` function `create_cat_var` takes in a `{metacore}` object, a reference variable - in this case `AGE` because that is the continuous variable `AGEGR1` is created from, and the name of the sub-grouped variable. It will take the controlled terminology from the sub-grouped variable and group the reference variables accordingly. 

Using a similar philosophy we can create the numeric version of `RACE` using the controlled terminology stored in the `{metacore}` object with the `create_var_from_codelist` function in `{metatools}`

```{r ct}
adsl_ct <- adsl_preds %>% 
   create_cat_var(metacore, ref_var = AGE, 
                  grp_var = AGEGR1, num_grp_var = AGEGR1N) %>% 
   create_var_from_codelist(metacore = metacore, 
                            input_var = RACE, 
                            out_var = RACEN)

head(adsl_ct, n=10)
```

Now we have sorted out what we can easily do with controlled terminology it is time to start deriving some variables.
Here you could refer directly to using the `{admiral}` template and [vignette](https://pharmaverse.github.io/admiral/articles/adsl.html) in practice, but for the purpose of this end-to-end ADaM vignette we will share a few exposure derivations from there.
We derive the start and end of treatment, the treatment duration, and the safety population flag.

```{r exposure}
adsl_raw <- adsl_ct %>%
  derive_vars_merged_dtm(
    dataset_add = admiral_ex,
    filter_add = (EXDOSE > 0 |
      (EXDOSE == 0 &
        str_detect(EXTRT, "PLACEBO"))) & nchar(EXSTDTC) >= 10,
    new_vars_prefix = "TRTS",
    dtc = EXSTDTC,
    order = vars(TRTSDTM, EXSEQ),
    mode = "first",
    by_vars = vars(STUDYID, USUBJID)
  ) %>%
  derive_vars_merged_dtm(
    dataset_add = admiral_ex,
    filter_add = (EXDOSE > 0 |
      (EXDOSE == 0 &
        str_detect(EXTRT, "PLACEBO"))) & nchar(EXENDTC) >= 10,
    new_vars_prefix = "TRTE",
    dtc = EXENDTC,
    time_imputation = "last",
    order = vars(TRTEDTM, EXSEQ),
    mode = "last",
    by_vars = vars(STUDYID, USUBJID)
  ) %>%
   derive_vars_dtm_to_dt(source_vars = vars(TRTSDTM, TRTEDTM)) %>%  #Convert Datetime variables to date 
   derive_var_trtdurd() %>% 
   derive_var_merged_exist_flag(
     dataset_add = admiral_ex,
     by_vars = vars(STUDYID, USUBJID),
     new_var = SAFFL,
     condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
   ) %>% 
   drop_unspec_vars(metacore) #This will drop any columns that aren't specificed in the metacore object

head(adsl_raw, n=10)
```

```{r, warning=FALSE, message=FALSE, include=FALSE}
# Create dummy variables to match metacore specs to avoid later errors
# In practice these would be mainly created using derivation functions from admiral
adsl_raw <- adsl_raw %>%
  mutate(
    SITEGR1 = NA,
    TRT01PN = NA,
    TRT01A = NA,
    TRT01AN = NA,
    AVGDD = NA,
    CUMDOSE = NA,
    ITTFL = NA,
    EFFFL = NA,
    COMP8FL = NA,
    COMP16FL = NA,
    COMP24FL = NA,
    DISCONFL = NA,
    DSRAEFL = NA,
    BMIBL = NA,
    BMIBLGR1 = NA,
    HEIGHTBL = NA,
    WEIGHTBL = NA,
    EDUCLVL = NA,
    DISONSDT = NA,
    DURDIS = NA,
    DURDSGR1 = NA,
    VISIT1DT = NA,
    VISNUMEN = NA,
    RFENDT = NA,
    DCDECOD = NA,
    EOSSTT = NA,
    DCSREAS = NA,
    MMSETOT = NA
  )
```

## Apply Metadata to Create an eSub XPT and Perform Associated Checks

Now we have all the variables defined we can run some checks before applying the necessary formatting. 
```{r checks, warning=FALSE, message=FALSE}

adsl_raw %>% 
   check_variables(metacore) %>% # Check all variables specified are present and no more
   check_ct_data(metacore) %>% # Checks all variables with CT only contain values within the CT
   order_cols(metacore) %>% # Orders the columns according to the spec
   sort_by_key(metacore) %>% # Sorts the rows by the sort keys 
   xportr_type(metacore) %>% # Coerce variable type to match spec
   xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata 
   xportr_label(metacore) %>% # Assigns variable label from metacore specifications 
   xportr_df_label(metacore) # Assigns dataset label from metacore specifications
```
