---
title: "Create Demographic table"
author: "Tadeusz Lewandowski"
date: "16/08/2022"
output: html_document
description: "An example of how to generate an Demogrphic output based on ADaM dataset"
---

## Introduction

This guide will show you how two pharmaverse packages, along with some from tidyverse, can be used to create a Demographic table as an end-to-end, using the Admiral ADSL data as input.

The two packages used with a brief description of their purpose are as follows:

* [`{rtables}`](https://roche.github.io/rtables/index.html): designed to create and display complex tables with R.
* [`{tern}`](https://insightsengineering.github.io/tern/main/index.html): contains analysis functions to create tables and graphs used for clinical trial reporting.

It is important to understand [`{rtables}`](https://roche.github.io/rtables/index.html) grammar by reading through the above linked package site, as these are fundamental to being able to use [`{tern}`](https://insightsengineering.github.io/tern/main/index.html). 

## Load Data and Required pharmaverse Packages

After installation of packages, the first step is to load our pharmaverse packages and input data.

```{r setup, message=FALSE, warning=FALSE, results='hold'}
options(repos = c(
  pharmaverse = 'https://pharmaverse.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))

library(admiral)
library(rtables)
library(tern)
library(dplyr)

# Read in input ADaM data 
data("admiral_adsl")

```
  
## Start preprocessing

The first step we are going to encode missing entries across groups of categorical variables in a data frame `adsl`.


```{r, error=TRUE}
adsl <- df_explicit_na(admiral_adsl)
```

Now we will change some and add new variable `AGEGRP`. In the new variable we also are going to add levels in order to keep the required sequence of new categories on the final output.

```{r}
adsl <- adsl %>%
  mutate(
    SEX = factor(case_when(
      SEX == "M" ~ "Male",
      SEX == "F" ~ "Female",
      SEX == "U" ~ "Unknown",
      SEX == "UNDIFFERENTIATED" ~ "Undifferentiated"
    )),
    AGEGRP = factor(case_when(
      between(AGE, 18, 40) ~ "18-40",
      between(AGE, 41, 64) ~ "41-64",
      AGE > 64 ~ ">=65"
    ),
    levels = c("18-40", "41-64", ">=65")
    )
  ) 

```

Now optionally we create lists: `vars` and `var_labels` which hold the variables and thier corresponding labels. We will use them in the next step.

```{r}
vars <- c("AGE", "AGEGRP", "SEX","RACE")

var_labels <- c(
  "Age (yr)",
  "Age groups",
  "Sex",
  "Race"
)
```

Now we create the demographic table using the [`{rtables}`](https://roche.github.io/rtables/index.html) and [`{tern}`](https://insightsengineering.github.io/tern/main/index.html) packages.

```{r}
result <- basic_table() %>%
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  summarize_vars(
    vars = vars,
    var_labels = var_labels
  ) %>%
  build_table(adsl)

result
```


