---
title: "Create ADSL"
author: "Christina Fillmore"
date: "03/02/2022"
output: html_document
description: "An example of how to generate an ADSL dataset from STDM"
---
The first step is to load our pharamverse libraries and data. Here we are trying to create an ADSL dataset using the pilot CDISC data.

```{r setup, warning=FALSE, results='hold'}
options(repos = c(
  pharmaverse = 'https://pharmaverse.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))

library(metacore)
library(metatools)
library(admiral)
library(xportr)
library(dplyr)
library(tidyr)
library(safetyData)
library(lubridate)
library(stringr)
# Read in data 
data("sdtm_dm")
data("sdtm_ds")
data("sdtm_ex")
data("sdtm_vs")
data("sdtm_lb")
data("sdtm_sv")
data("sdtm_sc")
data("sdtm_mh")
data("sdtm_qs")
# Read in metacore object 
load(metacore_example("pilot_ADaM.rda"))
metacore <- metacore %>% 
   select_dataset("ADSL")
```

## Start Building 


The first thing we are going to do is pull through all the columns that come directly from the SDTM datasets. You might know which datasets you are going to pull from directly already, but if you don't you can call `build_from_derived` with just an empty list and the error will tell you which datasets you need to supply. 

```{r, error=TRUE}
build_from_derived(metacore, list(), predecessor_only = FALSE)
```

In this case all the columns come from `DM` so that is the only dataset I will pass into `build_from_derived`. The resulting dataset has all the columns combined and any columns that needed renaming between SDTM and ADAM are renamed. 
```{r demographcis}
adsl_preds <- build_from_derived(metacore, 
                                 ds_list = list("dm" = sdtm_dm), 
                                 predecessor_only = FALSE, keep = TRUE)
head(adsl_preds)
```


Now we have the base dataset, we can start to create some variables. I am going to start with creating the subgroups using the control terminology, in this case AGEGR1. The metacore object holds all the metadata needed to make ADSL. Part of that metadat is the control terminology, which can help automate the creation of subgroups. We can look into the metacore object and see the control terminology for AGRGR1. 
```{r}
get_control_term(metacore, variable = AGEGR1)
```

Because this control terminology is written in a fairly standard format we can automate the creation of AGEGR1. The {metatools} function `create_cat_var` takes in a metacore object, a reference variable, in this case AGE because the is the continues variable AGEGR1 created from, and the name of the sub-grouped variable. It will take the control terminology from the sub-grouped variable and cut the reference variables accordingly. 

Using a similar philosophy we can create the numeric version of RACE and TRT01P using the control terminology stored in the metacore object with the `create_var_from_codelist` function in {metatools}

```{r ct}
adsl_preds %>% 
   create_cat_var(metacore, ref_var = AGE, 
                  grp_var = AGEGR1, num_grp_var = AGEGR1N) %>% 
   select(AGE, AGEGR1, AGEGR1N)

adsl_ct <- adsl_preds %>% 
   create_cat_var(metacore, ref_var = AGE, 
                  grp_var = AGEGR1, num_grp_var = AGEGR1N) %>% 
   create_var_from_codelist(metacore, RACE, RACEN) %>% 
   create_var_from_codelist(metacore, TRT01P, TRT01PN) %>% 
   mutate(
      SITEID = as.character(SITEID),
      SITEGR1 = SITEID, 
      TRT01A = TRT01P,
      TRT01AN = TRT01PN,
      ITTFL = if_else(!is.na(ARM) & ARM != "Screen Failure", "Y", "N"))
```


Now we have sorted out what we can we easily do with control terminology it is time to start coding up some variables. We are going to start with the exposure variables. We need to get the start and end of treatment as well as the treatment duration, the cumulative and the average dose. 

```{r exposure}
adsl_dates <- adsl_ct %>%
   derive_var_trtsdtm(dataset_ex = sdtm_ex) %>% # Derive Datetime of First Exposure to Treatment
   derive_var_trtedtm(dataset_ex = sdtm_ex) %>% # Derive Datetime of Last Exposure to Treatment
   derive_vars_dtm_to_dt(source_vars = vars(TRTSDTM, TRTEDTM)) %>%  #Convert Datetime variables to date 
   derive_var_trtdurd() %>% 
   mutate(TRTDURD = as.numeric(TRTEDT - TRTSDT) + 1, #derive_var_trtdurd
          CUMDOSE = TRT01PN * TRTDURD, # switch to admiral 
          AVGDD = CUMDOSE/TRTDURD,
          SAFFL = if_else(ITTFL == "Y" & !is.na(TRTSDT), "Y", "N")) %>% 
   drop_unspec_vars(metacore)
```



```{r disposition}
adsl_dispo <- adsl_dates %>% 
   # Derive a Disposition Status 
   derive_var_disposition_status(
      dataset_ds = sdtm_ds,
      new_var = EOSSTT,
      status_var = DSDECOD,
      filter = DSCAT == "DISPOSITION EVENT"
   ) %>%
   # Derive a Disposition Reason
   derive_vars_disposition_reason(
      dataset_ds = sdtm_ds,
      new_var = DCSREAS,
      reason_var = DSDECOD,
      filter = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE"
   ) %>% 
   # Derived Disposition Date 
   derive_var_disposition_dt(
      dataset_ds = sdtm_ds,
      new_var = RFENDT,
      dtc = DSSTDTC,
      filter_ds = DSCAT == "OTHER EVENT" & DSDECOD == "FINAL RETRIEVAL VISIT"
   ) %>% 
   # Getting the standardize dispositions from codelist 
   create_var_from_codelist(metacore, EOSSTT, DCDECOD) %>% 
   # Creating disposition flags 
   mutate(
      DISCONFL = if_else(DCDECOD != "COMPLETED", "Y", NA_character_),
      DSRAEFL = if_else(DCDECOD != "ADVERSE EVENT", "Y", NA_character_))

# Get end of treatment visit 
adsl_dispo <- sdtm_ds %>% 
   filter(DSTERM =='PROTOCOL COMPLETED') %>% 
   mutate(VISNUMEN = if_else(VISITNUM == 13, 12, VISITNUM)) %>% 
   select(USUBJID, VISNUMEN) %>% 
   left_join(adsl_dispo, ., by = "USUBJID")

```



```{r BMI}
# Get heights at screening cause those are the only heights available 
heights <- sdtm_vs %>% 
   filter(VISITNUM == 1, VSTESTCD == 'HEIGHT') %>% 
   select(USUBJID, VSTEST, VSSTRESN)

bmis <- sdtm_vs %>% 
   # Get baseline weight 
   filter( VISITNUM == 3, VSTESTCD == "WEIGHT") %>% 
   select(USUBJID, VSTEST, VSSTRESN) %>% 
   # Combine with height 
   bind_rows(heights) %>% 
   # Pivot to a row per subject and calculate BMIBL
   pivot_wider(names_from = VSTEST, values_from = VSSTRESN) %>% 
   mutate(BMIBL = compute_bmi(Height, Weight)) %>% 
   rename(WEIGHTBL = Weight, HEIGHTBL = Height) %>%
   # Create the BMI grouping using the control terminology 
   create_cat_var(metacore, BMIBL, BMIBLGR1)
bmis 

adsl_bmi <- left_join(adsl_dispo, bmis, by = "USUBJID")

```


## Apply Codelists

```{r Complete Flags}

completer_cal <- function(.data, wk_num, sv){
   new_col <- paste0("COMP", wk_num, "FL") %>% sym()

   sv %>% 
      group_by(USUBJID) %>% 
      filter(VISIT == paste("WEEK", wk_num)) %>% 
      select(USUBJID, SVSTDTC) %>% 
      left_join(.data, . , by = "USUBJID") %>% 
      mutate({{new_col }}:= ifelse(!is.na(SVSTDTC) && TRTEDT >= SVSTDTC, "Y", "N")) %>% 
      select(-SVSTDTC)
}

first_visit <- sdtm_sv %>% 
   filter(VISITNUM == 1) %>% 
   mutate(VISIT1DT = as_date(SVSTDTC)) %>%
   select(USUBJID, VISIT1DT) 

adsl_fls <- adsl_bmi %>% 
   completer_cal(8, sdtm_sv) %>% 
   completer_cal(16, sdtm_sv) %>% 
   completer_cal(24, sdtm_sv) %>% 
   left_join(first_visit, by= "USUBJID")
```

```{r Medical History}
adsl_ed <- sdtm_sc %>% 
   filter(SCTESTCD == "EDLEVEL") %>% 
   select(USUBJID, SCSTRESN) %>% 
   rename(EDUCLVL = SCSTRESN) %>% 
   left_join(adsl_fls, ., by = "USUBJID")

adsl_alz <- sdtm_qs %>% 
   group_by(USUBJID) %>% 
   filter(QSCAT == "ALZHEIMER'S DISEASE ASSESSMENT SCALE") %>%
   summarise(MMSETOT = sum(as.numeric(QSORRES), na.rm = TRUE)) %>% 
   left_join(adsl_ed, ., by = "USUBJID")

adsl_mh <- sdtm_mh %>% 
   filter(MHCAT == "PRIMARY DIAGNOSIS") %>% 
   mutate(DISONSDT = as_date(MHSTDTC)) %>%
   select(USUBJID, DISONSDT) %>% 
   left_join(adsl_alz, ., by = "USUBJID") %>% 
   mutate(DURDIS = interval(DISONSDT, VISIT1DT) %/% months(1)) %>% 
   create_cat_var(metacore, DURDIS, DURDSGR1)
```


```{r Efficacy Flag}
adsl_raw <- sdtm_qs %>% 
   filter(VISITNUM > 3) %>% 
   group_by(USUBJID) %>% 
   summarise(efffl = any(QSTEST == "ADAS-COG(11) Subscore") & any(QSTESTCD == "CIBIC")) %>% 
   left_join(adsl_mh, by = "USUBJID") %>% 
   mutate(EFFFL = if_else(efffl & SAFFL == "Y", "Y", "N")) %>% 
   drop_unspec_vars(metacore) #This will drop any columns that aren't specificed in the metacore object

```


Now we have all the variables defined we can run some checks before applying the necessary formatting. 
```{r checks}
test <- metacore$var_spec %>% 
   select(variable, type) %>% 
   mutate(dataset = "ADSL")

adsl_raw %>% 
   check_variables(metacore) %>% # Check all variables specified are present and no more
   check_ct_data(metacore) %>% # Checks all variables with CT only contain values within the CT
   order_cols(metacore) %>% # Orders the columns according to the spec
   sort_by_key(metacore) %>% # Sorts the rows by the sort keys 
   xportr_type(test) %>% # Coerce variable type to match spec
   xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata 
   xportr_label(metacore) %>% # Assigns variable label from metacore specifications 
   xportr_df_label(metacore) # Assigns dataset label from metacore specifications
```

